Gaia's constraints on the extremely massive ($\sim$$150\,M_\odot$) runaway star candidate VFTS682

### 2018-06-11 VFTS682_APJL.tex
selma minor things

\begin{table}
  \begin{center}
    \caption{Kinematics of VFTS682. }
    \begin{tabular}{llc|c|c}
      \hline
      \hline
      &Parameter & Units & Value & Ref.\\

       \hline
           \multicolumn{5}{l}{\emph{Gaia's proper motion for VFTS 682
      and average of the region}} \\
      \hline
          &$\mu_\mathrm{RA}$&[$\masyr$] & $1.843\pm 0.070$
                     & (1) \\        
               &$\mu_\mathrm{DEC}$&[$\masyr$] & $0.786\pm 0.080$
                     &  (1) \\        
                 & Correlation  ($\mu_\mathrm{RA}$, $\mu_\mathrm{DEC}$) &[$\masyr$] & $0.023$
                        & (1)  \\         
     %  \multicolumn{5}{l}{\emph{Gaia's average proper motion in the region}} \\         
       &$\langle\mu_\mathrm{RA}\rangle_\mathrm{R136}$&[$\masyr$] & $1.74\pm0.01$
                        & (2) \\
      &$\langle\mu_\mathrm{DEC}\rangle_\mathrm{R136}$&[$\masyr$]
                & $0.70\pm0.02$ &  (2)\\
\hline

@@

    %  &RA &[degree] &  \phantom{-}84.73 % $\pm$  0.03
   %                     & \multirow{2}{*}{\cite{evans:11}}\\[5pt]
    %  &DEC & [degree] & -69.07 % $\pm$  0.05
     %                   & \\[5pt]
      \multicolumn{5}{l}{\emph{Gaia's relative proper motion for VFTS
      682 }} \\
      \hline
      &$\delta\mu_\mathrm{RA}$  &[$\masyr$] & 0.103$\pm$0.080 & (1,4) \\
      &$\delta\mu_\mathrm{DEC}$  &[$\masyr$] & 0.086$\pm$0.100 &  (1,4) \\
      &$\delta\mu_\mathrm{}$  &[$\masyr$] & \SdM{0.13}$\pm$\SdM{0.09} &  (1,4) \\
                 &$v_\mathrm{2d}$  &[$\kms$] & \SdM{32}$\pm$\SdM{21} & (1,4)\\  
                     &$\theta$  &[degrees] &  14.2$\pm$\SdM{xxx}  & (1,4)\\  
                         &$v_\mathrm{2d, \parallel}$  &[$\kms$] & \SdM{xxx}$\pm$\SdM{xxx} & (1,4)\\  
 %

 \hline     
          \multicolumn{5}{l}{\emph{HST's independent determination of
      the relative proper motion for VFTS 682}} \\
            \hline
      &$\delta\mu_\mathrm{RA, HST}$  &[$\masyr$] & 0.01$\pm$0.13 & (3) \\
      &$\delta\mu_\mathrm{DEC, HST}$  &[$\masyr$] & 0.20$\pm$0.10 &  (3) \\
       &$\delta\mu_\mathrm{HST}$  &[$\masyr$] & 0.20$\pm$0.10 &  (3) \\
                  &$v_\mathrm{2d, HST}$  &[$\kms$] & \SdM{$47\pm24$} & (3)\\  
                      &$\theta^{*}$  &[degrees] &  \SdM{xxx}$\pm$\SdM{xxx}  & (1,4)\\  
                          &$v_\mathrm{2d, \parallel}$  &[$\kms$] & \SdM{xxx}$\pm$\SdM{xxx} & (1,4)\\  
 %

@@

    \end{tabular}
    \tablecomments
    { Proper motions are based on Gaia data unless indicated otherwise in subscript. The angle $\theta$ is defined such that $\theta=0$ corresponds to the direction pointing away from R136.
    (1) \cite{brown:18}, 
    (2) \cite{lennon:18}, 
    (3) \cite{platais:18},  
    (4) {\color{blue}This work}
    %The error on the RA and DEC positions, are of order $\sim$$10^{-2}\masyr$ in \emph{Gaia} DR2. %REPETITION: %  We adopt the peculiar
      % radial velocity calculated using the  NV $\lambda4944$ line, which is
      % less sensitive to the wind velocity structure (see also \Secref{data:vfts683}).
    }
  \end{center}
  \label{tab:vfts682}
\end{table}

\begin{figure*}[htbp]
  \centering
  \includegraphics[width=0.35\textwidth]{figures/dist_mu_region.pdf}
  \includegraphics[width=0.315\textwidth]{figures/angle_dist}
  \caption{Left panel: distribution of OB-type and Wolf-Rayet stars in proper
    motion relative to R136. VFTS682 is not an outlier, but
    its relative proper motion matches the expected value if it were indeed
    ejected from R136 assuming an age of $1.0\pm0.2$\,Myr. The top axis shows the conversion to physical units
    assuming a distance of 50\,kpc. Right panel:  distribution of
    angles between the relative proper motion direction and the radial
    direction to the star. The error bars for VFTS682 are large, but
    the best value is in agreement with the hypothesis of dynamical
    ejection. In both
    panels, the dark blue histograms contain 317 
    stars with error smaller than $0.1\,\mathrm{mas \
      yr^{-1}}\simeq25\,\mathrm{km\ s^{-1}}$ at 50\,kpc, the
    lighter blue histograms contain 36 stars with errors smaller than $0.05\,\mathrm{mas \
      yr^{-1}}$. \todo{ Todo: [1] insert correct errorbars in right pabel, [2] make linewidth, shading and figure size of both panels consistent [3] fix normalization to be consistent in both pabels}}
  \label{fig:mu_dist}
\end{figure*}

@@

The distance to 30 Doradus is $50$\,kpc  \citep[][]{pietrzynski:13}. The error on the distance determination is small ($\lesssim2\%$) and the possible difference between VFTS682 and the average of the region is probably even smaller (likely $\ll 1\%$). These uncertainties are negligible compared to the errors in the proper motion discussed below.

@@

VFTS682 is identified with the source id 4657685637907503744 in the
\emph{Gaia} DR2 catalog  as a 15.65 mag star in the G band
\citep{gaia:16,brown:18}.   The number of visibility periods,
i.e. groups of observations separated from other groups by a gap of at
least 4 days, used in the astrometric solution is 17 for this
star. The reported astrometric excess noise is zero.  These values
suggest that the \emph{Gaia} DR2  data for VFTS682 are
reliable. Generally, the Gaia errors are correlated
\citep[][]{lindengren:18}, but for this star the correlation coefficient is small, see Tab.~\ref{tab:vfts682}
. We can therefore treat the proper motion components as uncorrelated.

@@

We list their data in \Tabref{tab:vfts682}.  The errorbars are comparable with those we obtain from \emph{Gaia}.  However the direction and angle are different.  
\SdM{Can we please get theta Gaia and theta HST in the table.}

@@

\subsection{Is it a runaway star?}
\label{sec:runaway}
We first address the question of whether VFTS682 is a typical star
from the kinematic point of view, or whether it is a runaway star with
a significantly large peculiar velocity compared to its surrounding population. The former is what should
be expected if it formed in the relative isolation that we observe today.% , in relative
% isolation from other massive stars.

% Using the stars selected as described in \Secref{data:gaia} (a
% subset is shown in blue in \Figref{fig:main}), we find averaged proper motion components of
% $\langle\mu_\mathrm{RA}\rangle = 1.7240\pm0.0002\masyr$ and
% $\langle\mu_\mathrm{DEC}\rangle = 0.6986\pm0.0003\,\mathrm{mas\
% yr^{-1}}$. These values are
% in good agreement with what is found by \cite{lennon:18}. However, we
% emphasize their sensitivity to the choice of stars adopted to define
% the reference frame.
Subtracting the mean proper motion components given by
\citet{lennon:18} from the
proper motion of VFTS682 (see \Tabref{tab:vfts682}), we obtain the
components of proper motion of the star relative to the surrounding region
$\delta\mu_\mathrm{RA}^{Gaia} = 0.10 \pm 0.07\masyr$
and $\delta\mu_\mathrm{DEC}^{Gaia} = 0.08
\pm 0.09\masyr$. These components result in a
two-dimensional relative proper motion of $\delta \mu^{Gaia}=0.13\pm 0.09\,\mathrm{mas\
  yr^{-1}}$.

@@

% \begin{figure}%[htbp]
%   \centering
%   \includegraphics[width=0.5\textwidth]{figures/figure_polar_682-1.pdf}
%   \caption{Polar plot of the relative proper motion components for the stars
%     defining our reference frame. VFTS682 is indicated by the plus sign, and
%     other notable outliers \citep[see][]{lennon:18} are also labeled. Concentric dashed-lines denote relative proper motions of
%     0.1, 0.2, and 0.4$\masyr$. Positive angle
%     indicate a tangential component pointing in the counterclockwise
%     direction, with 0 degrees corresponding to proper motion pointing
%     radially outward from R136. The proper motion for VFTS682
%     corresponds to almost radial motion away from the cluster.
%   }
%   \label{fig:pm_polar}
% \end{figure}

The proper motion components can be converted into
the components of the relative transverse velocity $\delta v_\mathrm{RA}^{Gaia}=24\pm19\,\kms$,
$\delta v_\mathrm{DEC}^{Gaia}=20\pm23\,\kms$, assuming a distance of
50\,kpc. These can be combined obtaining a projected two-dimensional
velocity of $31\pm21\kms$. % (we do not account for the uncertainty in the distance
% estimate when propagating errors).
The radial velocity from
\cite{bestenlehner:11} then gives the third component along
the line of sight, % allowing us to calculate the three-dimensional
% peculiar speed of the star:
which added in quadrature to the transverse components results in a
three-dimensional velocity of $44 \pm 21\kms$.
% \begin{equation}
%   \label{eq:speed_around_Gaia}
%   v_\mathrm{pec}^{Gaia} = \sqrt{\left(\delta v_\mathrm{RA}^{Gaia}\right)^2
%     +\left(\delta v_\mathrm{DEC}^{Gaia}\right)^2+\left(\delta
%       v_\mathrm{rad}^{VFTS}\right)^2} = 44 \pm 21
%   \kms \ .
% \end{equation}

Therefore, two completely independent measures of the proper motion of
VFTS682 relative to the surrounding field, one from HST and one from
\emph{Gaia} DR2, yield values of the peculiar three-dimensional
velocity of VFTS682 % (\Eqref{eq:speed_around_HST} and
% \Eqref{eq:speed_around_Gaia}, respectively)
which would make it the most massive runaway star
known to date. However, the large errors on both the proper motion measures
require confirmation with future astrometric data.

@@

\begin{figure}%[tbp]
  \centering
  \includegraphics[width=0.48\textwidth]{./figures/main_plot_good_gaia_only}  
  \caption{The red solid arrow indicates the proper motion of VFTS682
    relative to the region from \emph{Gaia} DR2, starting from the present day position of
    the star. The yellow arrows indicate the possible
    directions of projected motion within the \emph{Gaia} DR2 errors, and are extended
    backwards (dashed) to illustrate the uncertainty on the origin of the
    star. The length of the prolongations is proportional to the relative proper motion
    times the age of VFTS682 \citep[$1.0\pm0.2$\,Myr,][]{schneider:18}.
    % \todo{load color picture on background? Fig with HST data
    % available, but confusing...}
  }

>>>>

\begin{table}
  \begin{center}
    \caption{Kinematics of VFTS682. }
    \begin{tabular}{llc|c|c}
      \hline
      \hline
      &Parameter & Units & Value & Ref.\\
      \hline
      \multicolumn{5}{l}{\emph{Coordinates}} \\
      \hline
         &$\mathrm{RA}_\mathrm{VFTS682}$&[degrees] & 84.7314 %$ \pm 0.036\,\mathrm{mas} $ 
                     & (1) \\        
               &$\mathrm{DEC}_\mathrm{VFTS682}$&[degrees] & -69.0741%$ \pm 0.048\,\mathrm{mas} $
                     & (1)  \\

                        &$\mathrm{RA}_\mathrm{R136}$&[degrees] & 84.6767
                     &  SIMBAD  \\        
               &$     \mathrm{DEC}_\mathrm{R136}$&[degrees] &  -69.1009
                     &  SIMBAD \\       
        &$      \delta\mathrm{RA}$  &[degrees] & 0.0547                      
        &  \SdM{your value? }
  \\        
               &$     \delta\mathrm{DEC}$  &[degrees] &0.0268 
                     &  \SdM{your value ? } \\

       \hline
           \multicolumn{5}{l}{\emph{Gaia's proper motion for VFTS 682
      and average of the region}} \\
      \hline
          &$\mu_\mathrm{RA}$&[$\masyr$] & $1.843\pm 0.070$
                     & (1) \\        
               &$\mu_\mathrm{DEC}$&[$\masyr$] & $0.786\pm 0.080$
                     &  (1) \\        
                 & $\rho\,(\mu_\mathrm{RA}, \mu_\mathrm{DEC})$ &[$\masyr$] & $0.023$
                        & (1)  \\         
     %  \multicolumn{5}{l}{\emph{Gaia's average proper motion in the region}} \\         
       &$\langle\mu_\mathrm{RA}\rangle_\mathrm{R136}$&[$\masyr$] & $1.74\pm0.01$
                        & (2) \\
      &$\langle\mu_\mathrm{DEC}\rangle_\mathrm{R136}$&[$\masyr$]
                & $0.70\pm0.02$ &  (2)\\
\hline

@@

      \multicolumn{5}{l}{\emph{Gaia's relative proper motion for VFTS
      682 }} \\
      \hline
      &$\delta\mu_\mathrm{RA}$  &[$\masyr$] & $0.103\pm0.080$ & (1,4) \\
      &$\delta\mu_\mathrm{DEC}$  &[$\masyr$] & $0.086\pm0.100$ &  (1,4) \\
      &$\delta\mu_\mathrm{}$  &[$\masyr$] & $0.13\pm0.09$ &  (1,4) \\
                 &$v_\mathrm{2d}$  &[$\kms$] & \SdM{$31\pm 21$} (not $32\pm21$!) & (1,4)\\  
                     &$\theta$  &[degrees] &  \SdM{$14.2_{-xx}^{+xx}$}  & (1,4)\\  
                         &$v_\mathrm{2d, \parallel}$  &[$\kms$] & $\SdM{xxx}\pm\SdM{xxx}$ & (1,4)\\  
 %

 \hline     
          \multicolumn{5}{l}{\emph{HST's relative proper motion for VFTS 682}} \\
            \hline
      &$\delta\mu_\mathrm{RA, HST}$  &[$\masyr$] & $0.01\pm0.13$ & (3) \\
      &$\delta\mu_\mathrm{DEC, HST}$  &[$\masyr$] & $0.20\pm0.10$ &  (3) \\
       &$\delta\mu_\mathrm{HST}$  &[$\masyr$] & $0.20\pm0.10$ &  (3) \\
                  &$v_\mathrm{2d, HST}$  &[$\kms$] & $47\pm24$ & (3)\\  
                      &$\theta^{*}$  &[degrees] &   \SdM{$xx_{-xx}^{+xx}$}   & (1,4)\\  
                          &$v_\mathrm{2d, \parallel}$  &[$\kms$] & $\SdM{xxx}\pm\SdM{xxx}$ & (1,4)\\  
 %

@@

    \end{tabular}
    \tablecomments
    { Proper motions are based on Gaia data unless indicated otherwise in subscript.    The conversion factor assuming a distance of 50\,kpc is such that $1\masyr$ corresponds to $237\kms$.  The correlation coefficient is indicated as $\rho$. The angle $\theta$ is defined such that $\theta=0$ for motions pointing away radially from R136.
    (1) \cite{brown:18}, 
    (2) \cite{lennon:18}, 
    (3) \cite{platais:18} and
    (4) {\color{blue}This work}.
    %The error on the RA and DEC positions, are of order $\sim$$10^{-2}\masyr$ in \emph{Gaia} DR2. %REPETITION: %  We adopt the peculiar
      % radial velocity calculated using the  NV $\lambda4944$ line, which is
      % less sensitive to the wind velocity structure (see also \Secref{data:vfts683}).
    }
  \end{center}
  \label{tab:vfts682}
\end{table}

\begin{figure*}[htbp]
  \centering
  \includegraphics[width=0.35\textwidth]{figures/dist_mu_region.pdf}
  \includegraphics[width=0.35\textwidth]{figures/angle_dist}
  \caption{Left panel: distribution of OB-type and Wolf-Rayet stars in proper
    motion relative to R136. VFTS682 is not an outlier, but
    its relative proper motion matches the expected value if it were indeed
    ejected from R136 assuming an age of $1.0\pm0.2$\,Myr. The top axis shows the conversion to physical units
    assuming a distance of 50\,kpc. Right panel:  distribution of
    angles between the relative proper motion direction and the radial
    direction to the star. The error bars for VFTS682 are large, but
    the best value is in agreement with the hypothesis of dynamical
    ejection. In both
    panels, the dark blue histograms contain 317 
    stars with error smaller than $0.1\,\mathrm{mas \
      yr^{-1}}\simeq25\,\mathrm{km\ s^{-1}}$ at 50\,kpc, the
    lighter blue histograms contain 36 stars with errors smaller than $0.05\,\mathrm{mas \
      yr^{-1}}$. \SdM{ Todo: [1] insert correct errorbars in right pabel, [2] make (a) linewidth, (b) line width of the hashed shading and (c)  figure height consistent. (You could try labeling the top x axis in the right panel in degrees.)  [3] fix normalization to be consistent in both pabels, [4] unit in axis label in righthand plot is not correct. Degrees but radian. [4] Legen: the rel error limits that we adopted are missing a unit. Milia arrsecond?    }}
  \label{fig:mu_dist}
\end{figure*}

@@

For the distance we adopt the distance to the LMC of $50$\,kpc. The error on the distance determination is small ($\lesssim2\%$, \citealt{pietrzynski:13}) and any possible offset between  R136 or VFTS682  and the distance we adopted for the LMC is probably much smaller (likely $\ll 1\%$, e.g. \citealt{Luks+1992}). These uncertainties are negligible compared to the errors in the proper motion discussed below.

@@

VFTS682 is identified with the source id 4657685637907503744 in the
\emph{Gaia} DR2 catalog  as a 15.65 mag star in the G band
\citep{gaia:16,brown:18}.   The number of visibility periods,
i.e. groups of observations separated from other groups by a gap of at
least four days, used in the astrometric solution is seventeen for this
star. The reported astrometric excess noise is zero.  These values
suggest that the \emph{Gaia} DR2  data for VFTS682 are
reliable. Gaia errors are correlated \citep[][]{lindengren:18}, but for this star the correlation coefficient is very small, $\rho\,(\mu_\mathrm{RA}, \mu_\mathrm{DEC})$  = 0.023, where zero is for completely un-correlated data and one is for fully correlated data. We can thus safely treat the proper motion components as uncorrelated.

@@

We list their data in \Tabref{tab:vfts682}.

%The errorbars are comparable with those we obtain from \emph{Gaia}.  However the direction and angle are different.

@@

\SdM{Since the errors for HST and Gaia are so similar and yield to different directions, I think we should probably restructure this paragraph.  I liked how you break down the story in two separate questions, but given Fig 1 I think it may not work well.  The first question should have adressed it in context of the motions of the other stars. And it has a very typical motion. I think that instead, you should simply describe Fig 1 step by step.  I think that will be stronger.  }

\subsection{Is it a runaway star?}
\label{sec:runaway}
We first address the question whether VFTS682 is a typical star
from the kinematic point of view (which is what would be expected if it formed in relative isolation), or whether has a significant peculiar velocity compared to its surrounding population (which is what we would expect if the star has a runaway origin).

After subtracting the mean proper motion of the field, we obtain the relative two-dimensional relative proper motion. We obtain  $\delta \mu_{Gaia}=0.13\pm 0.09\masyr (32\pm 21\kms)$   and $\delta \mu_{HST}=0.20\pm 0.10\masyr (47 \pm 24\kms )$.

Both indicate a non-zero peculiar 2D motion that would make it a slow runaway star, but in both cases the meaurements are also consistent with zero within 2 sigma.

Figure~\ref{fig:mu_dist} shows the distribution of relative proper motions for other VFTS O, B and WR stars subselecting either those with relative proper motion errors smaller than $0.1\masyr$ (yielding 317 stars) or a more restrictive cut including only those with proper motion errors smaller than $0.05\masyr$ (yielding 36) stars.   We see that the derived Gaia proper motion is very typical for stars in the field.

\SdM{XXX We need to rethink }

%Therefore, two completely independent measures of the proper motion of VFTS682 relative to the surrounding field, one from HST and one from \emph{Gaia} DR2, yield values of the peculiar three-dimensional velocity of VFTS682 % (\Eqref{eq:speed_around_HST} and
% \Eqref{eq:speed_around_Gaia}, respectively)
%which would make it the most massive runaway star known to date. However, the large errors on both the proper motion measures require confirmation with future astrometric data.

.

%Subtracting the mean proper motion components given by \citet{lennon:18} from the proper motion of VFTS682 (see \Tabref{tab:vfts682}), we obtain the components of proper motion of the star relative to the surrounding region $\delta\mu_\mathrm{RA}^{Gaia} = 0.10 \pm 0.07\masyr$ and $\delta\mu_\mathrm{DEC}^{Gaia} = 0.08 \pm 0.09\masyr$. These components result in a two-dimensional relative proper motion of $\delta \mu^{Gaia}=0.13\pm 0.09\,\mathrm{mas\ yr^{-1}}$.

@@

%The proper motion components can be converted into the components of the relative transverse velocity $\delta v_\mathrm{RA}^{Gaia}=24\pm19\,\kms$, $\delta v_\mathrm{DEC}^{Gaia}=20\pm23\,\kms$, assuming a distance of 50\,kpc. These can be combined obtaining a projected two-dimensional velocity of $31\pm21\kms$.

%The radial velocity from \cite{bestenlehner:11} then gives the third component along the line of sight,  which added in quadrature to the transverse components results in a three-dimensional velocity of $44 \pm 21\kms$.
% \begin{equation}
%   \label{eq:speed_around_Gaia}
%   v_\mathrm{pec}^{Gaia} = \sqrt{\left(\delta v_\mathrm{RA}^{Gaia}\right)^2
%     +\left(\delta v_\mathrm{DEC}^{Gaia}\right)^2+\left(\delta
%       v_\mathrm{rad}^{VFTS}\right)^2} = 44 \pm 21
%   \kms \ .
% \end{equation}

@@

\begin{figure}%[tbp]
  \centering
  \includegraphics[width=0.48\textwidth]{./figures/main_plot_good_gaia_only}  
  \caption{\SdM{I don't see a reason why we would not show the HST data, except for wanting to hide it maybe.  Let's sit together and find a good way to plot both.  (Hope you still have the script to show the HST direction. )}  The red solid arrow indicates the proper motion of VFTS682. 
    relative to the region from \emph{Gaia} DR2, starting from the present day position of
    the star. The yellow arrows indicate the possible
    directions of projected motion within the \emph{Gaia} DR2 errors, and are extended
    backwards (dashed) to illustrate the uncertainty on the origin of the
    star. The length of the prolongations is proportional to the relative proper motion
    times the age of VFTS682 \citep[$1.0\pm0.2$\,Myr,][]{schneider:18}.
    % \todo{load color picture on background? Fig with HST data
    % available, but confusing...}
  }
